Formalization of Typed Lua

### 2015-06-09 doc/papers/paper/paper.tex
another rewrite of related work section of DLS paper, for consolidation and simplification

Typed Clojure~\cite{bonnaire-sergeant2012typed-clojure} is an
optional type system for Clojure.
The type system of Typed Clojure includes polymorphic function types,
union types, intersection types, lists, vectors, maps, sets, and recursive types.
Function types can have {\em rest} parameters, which are similar
to our variadic types, but Typed Clojure functions cannot return
multiple results. Typed Clojure also uses occurrence typing

Dart~\cite{dart} is class-based object-oriented programming
language with an optional type system. Dart's type system
is nominal, and includes polymorphic types. The type
system is unsound by design, with instances of polymorphic
types being covariant in their type arguments.

TypeScript~\cite{typescript} is a JavaScript extension
that includes optional type annotations and class-based
object-oriented programming.
The type system of TypeScript is structural, based
on subtyping, and has polymorphic types, union types,
function types, array types, tuple types, recursive types,
record types, and
object types. It is unsound by design, although there is a
sound subset that forms the basis of a gradual type
system~\cite{bierman2014typescript}. Although JavaScript
also has the same idiom of building a record through
imperative assignment to an initially empty record,
TypeScript does not have evolution of its record and
object types; the fact that {\tt undefined} is a bottom 
type in its subtyping relation still lets programmers use
the idiom, at the cost of type safety.

TeJaS~\cite{lerner2013tejas} is a framework for the
construction of different type systems for JavaScript.
It has a base type system with extensible typing rules.
As an example, the paper shows how TeJaS can simulate the
type system of TypeScript.

Politz et al.~\cite{politz2012semantics} proposes semantics
and types for objects with first-class member names, a well-known
feature from scripting languages. Their type system uses string 
patterns to describe the members of
an object, and define a complex subtyping relation to validate
these patterns.
They also provide an implementation of their system to JavaScript.

Gradualtalk~\cite{allende2013gts} is a Smalltalk dialect that
with a gradual type system. Its type system combines
 nominal and structural typing.
It includes function types, union types, structural object types,
nominal object types types, a self type, and parametric polymorphism. Like Typed Lua, Gradualtalk formalizes the
interaction between typed and untyped code with consistent-subtyping.

Gradualtalk inserts runtime checks that ensure dynamically
typed code does not violate statically typed code.
Allende et al.~\cite{allende2013cis} performs a careful
evaluation of the performance impact of these checks, and
conclude that the performance hit is large enough that
Gradualtalk needs to have an option that turns off these
checks, downgrading Gradualtalk into an optional type
system.

@@

Mypy~\cite{mypy} is an optional type system for Python.
The type system of mypy is similar to the type system of
Reticulated Python, but mypy has parametric polymorphism and
does not insert runtime checks.

Guido van Rossum, Python's designer, proposed a
standard syntax for type annotations in Python
that is inspired by mypy~\cite{PEP484}.
The goal of this proposal is to make it easier to build
static analysis tools for Python.
Typing~\cite{typing} is a tool that is being developed to
implement this proposal.

Hack~\cite{hack} is an optionally-typed dialect of PHP
that runs on the Hip Hop Virtual Machine (HHVM).
It has nominal object types, parametric polymorphism,
nullable types, array types, function types, and tuple
types.

The Ruby Type Checker~\cite{ren2013rtc} is a library that
augments the regular dynamic type checking of the Ruby
implementation with extra runtime checks generated from type
annotations that the programmer can use on classes and
methods. It has nominal object types, union types, intersection
types, method types, and parametric polymorphism.

Grace~\cite{black2013sg} is an object-oriented language
with optional typing. Like Dart, it is a language
that has been designed from scratch to have an
optional type system, instead of an existing dynamically
typed language that has been retrofitted with a type
system. Unlike Dart, the type system of Grace is not
deliberately unsound.

>>>>

Gradualtalk~\cite{allende2013gts} is a Smalltalk dialect 
with a gradual type system. Its type system combines
nominal and structural typing.
It includes function types, union types, structural object types,
nominal object types types, a self type, and parametric polymorphism. Like Typed Lua, Gradualtalk formalizes the
interaction between typed and untyped code with consistent-subtyping. Due to the performance impact of
the runtime checks that ensure the gradual typing guarantees,
Gradualtalk can be downgraded into an optional type system
through a configuration switch~\cite{allende2013cis}.

@@

Several dynamically typed languages now have optional
type systems: Clojure~\cite{bonnaire-sergeant2012typed-clojure},
JavaScript~\cite{typescript}, Python~\cite{mypy}, 
and PHP~\cite{hack}. While Lua has some similarities to
all of these languages, none of these optional type
systems have the features described in this paper.
Most of these languages do not have idioms
that inspired these features; JavaScript has the
idiom of adding fields to an initially empty
record through assignment, but TypeScript sidesteps
the issue by having allowing an empty record to
have any record or object type, as the type of
missing fields is the bottom type of its type system.

Dart~\cite{dart} and Grace~\cite{black2013sg} are
two languages that have been designed from scratch
to have an optional type system, instead of being
existing dynamically typed languages with a retrofitted 
type system.
